---
title: "R Notebook"
output: html_notebook
---


My set-up chunk:
```{r setup}
library(readxl) #for reading excel
```

Read in the excel file of differences. This gives us T1-T2 values for each coordinate for each landmark as well as the euclidean distance of T1 and T2 for each landmark.
```{r read_in_data}
angles <- as.data.frame(read_excel("angles.xlsx"))
```

This extracts the euclidean distances along with the patient identifiers.
```{r just_distances}
numcolmax <- length(t1_minus_t2[1, ]) #original number of columns

#str(angles) #look at; I commented this out using "#"; you can remove this

```
We have 26 patients and 6 angles; there is no missing data.

This will be a basic exploratory analysis of angles.
```{r eda}
as.data.frame(cbind(minimum = sapply(angles,min, na.rm = TRUE), maximum =  sapply(angles,max, na.rm = TRUE), average = sapply(angles,mean, na.rm = TRUE), median = sapply(angles,median, na.rm = TRUE))) #summary statistics

for(angle in 2:7){
  plot(cos(angles[, angle]/180*pi), as.numeric(sin(angles[, angle]/180*pi)), ylim = c(-1,1), xlim = c(-1,1))
} #here I plot each angle on the unit circle
```

As each angle is relatively close to each other, this suggests that we can treat the angles as numbers.


Now let's look at asymmetries
```{r compute_asymmetries}
asym_angles <- angles[ , 1]
for (landmark in 1:3) {
   asym_angles <- as.data.frame(cbind(asym_angles, as.vector(abs(angles[, (2*landmark)] - angles[, (2*landmark + 1)]))))
}

names(asym_angles) <- names(angles[ , c(1, seq(2, 6, by = 2))])
```

This will be a basic exploratory analysis of asymmetries.
```{r eda_asymmetries}
as.data.frame(cbind(minimum = sapply(asym_angles, min, na.rm = TRUE), maximum =  sapply(asym_angles, max, na.rm = TRUE), average = sapply(asym_angles, mean, na.rm = TRUE), median = sapply(asym_angles, median, na.rm = TRUE))) #summary statistics
```

Now, note that the sample average of asymmetries for all sets of angles is not very large. On average, we have asymmetries of 1.4, 4.5, and 5.4 degrees.

