---
title: "R Notebook"
output: html_notebook
---


My set-up chunk:
```{r setup}
library(readxl) #for reading excel
```

Read in the excel file of differences. This gives us T1-T2 values for each coordinate for each landmark as well as the euclidean distance of T1 and T2 for each landmark.
```{r read_in_data}
t1_minus_t2 <- read_excel("t1_minus_t2.xlsx")
```

This extracts the euclidean distances along with the patient identifiers.
```{r just_distances}
numcolmax <- length(t1_minus_t2[1, ]) #original number of columns
distances <- as.data.frame(t1_minus_t2[ , c(1, seq(5, numcolmax, by = 4))]) #extracting, turning into dataframe

#str(distances) #look at; I commented this out using "#"; you can remove this

#columns were not in order (left and right were not necessarily beside each other); fixed using sort

distances <- distances[ , c(1, order(names(distances[, 2:45])) + 1)] #sorting, not including the first column

#str(distances) #look at; manually sorting it is; I commented this out using "#"; you can remove this

distances_sorted <- distances[ , c(1:20, 22, 21, 23, 24, 27, 25, 28, 26, 29, 32, 35, 30, 33, 31, 34, 36, 40, 37, 41, 38, 42, 39, 43, 44, 45)]  #here I am sorting manually; this could be rearranged in excel as well before starting

str(distances_sorted) #look at; it looks good now

```
We have 26 patients and 44 landmarks; there is missing data (coded as NA). It is not in order; so I sorted it

This will be a basic exploratory analysis of distances.
```{r eda}
as.data.frame(cbind(minimum = sapply(distances_sorted,min, na.rm = TRUE), maximum =  sapply(distances_sorted,max, na.rm = TRUE), average = sapply(distances_sorted,mean, na.rm = TRUE), median = sapply(distances_sorted,median, na.rm = TRUE))) #summary statistics
```

Note that the median < mean for most landmarks suggesting a right skew (right tail)

Now let's look at asymmetries
```{r compute_asymmetries}
asym <- distances_sorted[ , 1]
for (landmark in 1:22){
   asym <- as.data.frame(cbind(asym, as.vector(abs(distances_sorted[, (2*landmark)] - distances_sorted[, (2*landmark + 1)]))))
}

names(asym) <- names(distances_sorted[ , c(1, seq(2, 44, by = 2))])
```

This will be a basic exploratory analysis of asymmetries.
```{r eda_asymmetries}
as.data.frame(cbind(minimum = sapply(asym, min, na.rm = TRUE), maximum =  sapply(asym, max, na.rm = TRUE), average = sapply(asym, mean, na.rm = TRUE), median = sapply(asym, median, na.rm = TRUE))) #summary statistics
```

Now, note that the sample average of asymmetries for many landmarks is not over 2mm; thus for each of these, we will not be able to say that the average asymmetry is significantly greater than 2mm. We are left with 2 landmarks with average asymmetry over 2mm; however, they seem to have some outliers (looking at the maximums). Is this a data entry error? Looking at the medians, I highly doubt we will find anything that is both significant and not driven by these outliers.

Because of this, I will most likely change the analysis to looking at 1mm difference because the choice of the 2mm threshold was arbitrary.

I want to take a closer look at the two landmarks with the outliers. Here is a boxplot for each.

```{r landmarks_with_outliers}

boxplot(asym$k_l_dist)

print(asym$k_l_dist)

boxplot(asym$u6_l_furc_dist)

print(asym$u6_l_furc_dist)

```

```{r}
#This below is no longer needed. This is where I manually removed the data with errors (the extreme outliers)

# distances_sorted$k_r_dist[22] <- NA #replacing extreme outlier with NA
# 
# distances_sorted$k_l_dist[22] <- NA #paired observation
# 
# distances_sorted$u6_l_furc_dist[19] <- NA #replacing extreme outlier with NA
# 
# distances_sorted$u6_r_furc_dist[19] <- NA #paired observation

```

```{r eda_NAs}
as.data.frame(cbind(minimum = sapply(distances_sorted,min, na.rm = TRUE), maximum =  sapply(distances_sorted,max, na.rm = TRUE), average = sapply(distances_sorted,mean, na.rm = TRUE), median = sapply(distances_sorted,median, na.rm = TRUE))) #summary statistics
```

Those two outliers are very extreme. My belief is that they are data-entry errors. 90mm for example is about 3 1/2 inches.

Now let's look at asymmetries
```{r compute_asymmetries_NAs}
asym <- distances_sorted[ , 1]
for (landmark in 1:22){
   asym <- as.data.frame(cbind(asym, as.vector(abs(distances_sorted[, (2*landmark)] - distances_sorted[, (2*landmark + 1)]))))
}

names(asym) <- names(distances_sorted[ , c(1, seq(2, 44, by = 2))])
```

This will be a basic exploratory analysis of asymmetries.
```{r eda_asymmetries_NAs}
as.data.frame(cbind(minimum = sapply(asym, min, na.rm = TRUE), maximum =  sapply(asym, max, na.rm = TRUE), average = sapply(asym, mean, na.rm = TRUE), median = sapply(asym, median, na.rm = TRUE))) #summary statistics
```



